---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# fax

<!-- badges: start -->
[![R-CMD-check](https://github.com/pedrobtz/fax/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pedrobtz/fax/actions/workflows/R-CMD-check.yaml)
[![live](https://github.com/pedrobtz/fax/actions/workflows/live.yaml/badge.svg)](https://github.com/pedrobtz/fax/actions/workflows/live.yaml)
<!-- badges: end -->

fax reports facts about where R is running and what it may actually use.

Inside a container, `parallel::detectCores()` returns the node's 64 cores while
the pod may use 4, and `/proc/meminfo` shows the node's memory while the
container is limited to 8 GiB. Thread pools sized on those numbers oversubscribe
the CPU, and memory budgets end in OOM kills. fax answers three questions on
bare metal, VMs, containers, Kubernetes pods and Azure, on Linux, macOS and
Windows:

1. **Where am I running?** OS, hardware, virtualization, container, Kubernetes,
   cloud, R and Python.
2. **What can I actually use?** Effective CPU cores and memory, next to the
   host's.
3. **What is installed?** System, R and Python packages.

## Installation

Install fax from CRAN:

``` r
install.packages("fax")
```

Or the development version from [GitHub](https://github.com/pedrobtz/fax):

``` r
# install.packages("pak")
pak::pak("pedrobtz/fax")
```

## Example

```{r example}
library(fax)
facts()
```

In a Kubernetes pod limited to 2 CPUs and 1 GiB on an 8-CPU, 32 GiB node:

```
#> <fax facts>
#> os       Ubuntu 22.04
#> runs in  Kubernetes pod default/web-6b8f9-2xkqz, containerd container
#> runs on  vm (unknown)
#> cpu      host 8 | effective 2
#> memory   host 32G | limit 1G | available 824M
```

Size worker pools with what the process may use:

``` r
future::plan(future::multisession, workers = effective_cores())
data.table::setDTthreads(effective_cores())
```

Add CPU and memory use to log lines, cheaply (about 170 µs on Linux):

``` r
message("batch done ", usage_line())
#> batch done cpu=1.8/4 thr=3% mem=2.1G/8G(26%) rss=1.2G
```

Every fact records its status and source, and fax never fails: what it cannot
determine is `NA`.

```{r}
head(facts_df()[, c("fact", "status", "source")])
```

fax makes no network requests unless you ask for cloud metadata with
`facts(cloud = TRUE)`, never reads credentials, and redacts secrets in
environment variables.
