---
title: "Getting started with fax"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fax}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# Building the vignette (e.g. in R CMD check) must never start Python.
options(fax.skip = "runtime.python")
```

fax reports facts about where R is running and what it may use. This article
shows the main functions on the machine that built it, then walks through a
container in Kubernetes, where the numbers fax gives you differ most from the
usual ones.

```{r}
library(fax)
```

## A snapshot

`facts()` returns a snapshot. Printing it gives a summary:

```{r}
f <- facts()
f
```

Facts are grouped in namespaces such as `os`, `cpu` and `memory`. Access a
namespace with `$`, a group or a single fact with `[[`:

```{r}
names(f)
f[["cpu.host"]]
f[["memory.effective.limit"]]
```

Nothing is read until you ask for it, and each fact is read once per session.
`facts(refresh = TRUE)` reads again. `fact()` gets one value directly:

```{r}
fact("cpu.effective")
```

The help page `?fax_facts` lists every fact.

## Status and source

A fact that cannot be determined is `NA`; fax never stops with an error.
`facts_df()` says why, and where each value came from:

```{r}
df <- facts_df(facts("cpu"))
rows <- df$fact %in% c("cpu.model", "cpu.affinity", "cpu.cgroup.quota", "cpu.effective")
df[rows, c("fact", "status", "source")]
```

`status` is `"ok"`, `"not_applicable"` (e.g. cgroup facts on macOS),
`"unavailable"` (the data is missing) or `"error"`. Use `strict = TRUE` to turn
errors into R errors while debugging.

## How many cores, how much memory?

`parallel::detectCores()` counts the host's CPUs. Inside a container the
process is usually allowed far fewer. `effective_cores()` takes the smallest
of:

* the host's logical CPUs,
* the process's CPU affinity,
* the cgroup's cpuset,
* the cgroup's CPU quota, rounded up (a 1.5-core quota can keep 2 threads busy).

```{r}
effective_cores()
effective_memory()
```

`effective_memory()` is the smaller of host memory and the cgroup memory
limit. `effective_memory("available")` is what can still be allocated, counting
reclaimable page cache as free.

Use them to size worker pools:

```{r, eval = FALSE}
future::plan(future::multisession, workers = effective_cores())
data.table::setDTthreads(effective_cores())
options(mc.cores = effective_cores())
```

Both ignore settings such as `mc.cores` or `MC_CORES` on purpose: they describe
the resources, not a policy. `parallelly::availableCores()` combines both.

## In a Kubernetes pod

Here is fax in a pod limited to 2 CPUs and 1 GiB on a node with 8 CPUs and 32
GiB. The limits are set on the pod's cgroup, a parent of the container's:

```{r, eval = FALSE}
facts()
#> <fax facts>
#> os       Ubuntu 22.04
#> runs in  Kubernetes pod default/web-6b8f9-2xkqz, containerd container
#> runs on  vm (unknown)
#> cpu      host 8 | effective 2
#> memory   host 32G | limit 1G | available 824M
#> Use facts_df() for every fact with its status and source.
```

`detectCores()` would say 8. The CPU facts show where 2 comes from:

```{r, eval = FALSE}
str(facts()$cpu[c("host", "affinity", "cgroup", "effective")])
#> List of 4
#>  $ host     :List of 3
#>   ..$ logical : int 8
#>   ..$ physical: int 4
#>   ..$ sockets : int 1
#>  $ affinity : int 8
#>  $ cgroup   :List of 3
#>   ..$ quota : num 2
#>   ..$ cpuset: int 8
#>   ..$ weight: num 79
#>  $ effective: int 2
```

For memory, the container has used 300 MiB, of which 100 MiB is inactive page
cache the kernel can reclaim. Like the kubelet, fax counts the remaining 200
MiB as the working set, leaving 824 MiB available.

Kubernetes facts come from the environment and from files the kubelet mounts.
The node name, pod IP and resources are only visible through the Downward API:
expose them as environment variables (fax looks for `NODE_NAME`, `POD_IP`,
`POD_UID`, `POD_NAME`, `CPU_LIMIT`, `MEMORY_LIMIT`, `CPU_REQUEST` and
`MEMORY_REQUEST`, or set `options(fax.k8s.env = )`) and mount labels and
annotations at `/etc/podinfo` (or set `options(fax.k8s.podinfo = )`). fax never
reads the service account token.

## Output formats

`facts_df()` gives a data frame, `as.list()` a nested list, and `facts_json()`
JSON (with the jsonlite package):

```{r, eval = requireNamespace("jsonlite", quietly = TRUE)}
facts_json(namespaces = "cgroup", pretty = TRUE)
```

## Options

| Option | Purpose |
|:--|:--|
| `fax.root` (or env `FAX_ROOT`) | Read `/proc`, `/sys` and `/etc` below another root, e.g. a host mounted at `/host`. |
| `fax.cloud` | Allow network requests to the cloud metadata service (same as `cloud = TRUE`). |
| `fax.strict` | Re-raise resolver errors. |
| `fax.skip` | Facts or namespaces to skip, e.g. `c("packages", "runtime.python")`. |
| `fax.redact`, `fax.redact_allowlist` | How `env.vars` hides secrets: `"default"`, `"allowlist"` or `"none"`. |
| `fax.k8s.env`, `fax.k8s.podinfo` | Downward API environment variable names and volume path. |

## Further reading

Articles on the package website cover:

* Azure: what fax detects offline and what instance metadata adds.
* Logging: `usage()` for CPU and memory in every log line.
* Inventory: installed system, R and Python packages.
