---
title: "Installed packages"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The `packages` namespace lists what is installed: system packages, R packages
and Python packages. It is opt-in, because the lists can be long: it is left out
of `print(facts())`, `as.list()` and `facts_df()` unless you ask for it.

```{r}
library(fax)
f <- facts("packages")
f[["packages.system.manager"]]
f[["packages.system.count"]]
head(f[["packages.r"]][, c("name", "version", "source")])
```

## Sources

| Fact | Read from |
|:--|:--|
| `packages.system.installed` | dpkg `/var/lib/dpkg/status` (or distroless `status.d/`), apk `/lib/apk/db/installed`, pacman `/var/lib/pacman/local`, one `rpm -qa` query, Homebrew's `Cellar` and `Caskroom` directories, or the Windows Uninstall registry keys. |
| `packages.r` | `installed.packages()`, with where each package came from (`Repository` or the remote type). |
| `packages.python` | `*.dist-info` and `*.egg-info` directories in the Python environment's site-packages, plus `conda-meta` in conda environments. |

fax never runs `pip`, `conda`, `brew` or `apt`, and never starts Python inside
R. Python is found through `RETICULATE_PYTHON`, `VIRTUAL_ENV`, `CONDA_PREFIX`
or the `PATH` (and the py launcher on Windows).

## An environment report

A typical use is attaching the environment to a bug report or a job log:

```{r, eval = FALSE}
namespaces <- c("os", "cpu", "memory", "runtime", "packages")
writeLines(facts_json(facts(namespaces), namespaces, pretty = TRUE), "environment.json")
```
