---
title: "CPU and memory in log lines"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`usage()` reads the current CPU and memory use of the R process and its
container. It is built to be called on every log line: everything static is
worked out once per session, and each call reads three small files on Linux
(about 170 microseconds).

```{r}
library(fax)
u <- usage()
u
unclass(u)
```

`usage_line()` gives the same as one string:

```{r, eval = FALSE}
message("batch 12 done ", usage_line())
#> batch 12 done cpu=1.8/4 thr=3% mem=2.1G/8G(26%) rss=1.2G
```

## Fields

| Field | Meaning |
|:--|:--|
| `cpu_process` | CPU cores used by this R process since the previous call. |
| `cpu_cgroup` | CPU cores used by the whole container since the previous call, including parallel workers. |
| `cpu_throttled` | Share of CPU scheduling periods in which the container was throttled. |
| `cpu_limit` | CPU cores the process may use. |
| `mem_rss` | Resident memory of this process. |
| `mem_cgroup` | Memory used by the container, including page cache. |
| `mem_limit` | The memory limit. |
| `mem_pct` | `mem_cgroup / mem_limit`, or `mem_rss` as a share of host memory outside containers. |

CPU values are rates, so they need two calls: the first reports the process's
average since it started, and the container fields start on the second call.
After `fork()` (for example in `parallel::mclapply()` workers) the baseline
starts again.

`cpu_throttled` is the number to watch in containers: a high value means the
container keeps hitting its CPU quota, which makes everything slower without
showing up as high CPU use.

On macOS and Windows `mem_rss` needs the ps package, and the container fields
are `NA`.

## With logging packages

fax depends on none of them; `usage_line()` is just a string.

```{r, eval = FALSE}
# logger
logger::log_info("model fitted {fax::usage_line()}")

# lgr: as a custom field, or the numbers themselves for JSON appenders
lgr::lgr$info("model fitted", usage = fax::usage_line())
lgr::lgr$info("model fitted", usage = as.list(unclass(fax::usage())))
```

In very hot loops, `max_age` returns the previous result if it is recent
enough:

```{r, eval = FALSE}
for (i in seq_len(1e6)) {
  if (i %% 1000 == 0) message(usage_line(max_age = 5))
}
```
